Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-11-03 doc/paper.tex
new stuff in paper

\begin{IEEEkeywords}
large language models, reinforcement learning, minigrid, proximal policy optimization
\end{IEEEkeywords}

@@

The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either interesting or boring. After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do. In \cite{idm}, another paper proposing using the LLM as a policy optimizer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed. Here, too, the LLM response is fed back into a task-specific RL model, which in turn influences policy probabilities of certain actions being taken. In general, many of the novel methods suggested involve the LLM suggesting actions for the agent to perform in a given context, and its output influencing the agent policy.

@@

We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy optimization ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***. All methods use the same hyperparameters for the PPO, which can be found in the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one. We do this eight times per method, and compare the average results against each other.

@@

We also notice a significant performance increase when using LLM policy optimization, even somewhat more than from reward shaping. Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM policy shaping. The standard deviation when using LLM policy optimization is also similarly large to PPO and reward shaping (NUMBER). By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.

***ADD PARAGRAPH ABOUT USING BOTH TOGETHER*** However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy optimization together. Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy optimization, with a similar standard deviation of NUMBER.

@@

Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER). LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping. Also here giving the highest average reward per episode, the combination of reward shaping and policy optimization gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy optimization.

\section{Conclusion}

It works man, dogski did it!

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so. This has been done both through the LLM shaping the RL reward function, and through the LLM influencing the action probabilities of the RL agent, as well as through combining these two methodologies.

For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.

>>>>

\begin{IEEEkeywords}
large language models, reinforcement learning, minigrid, proximal policy influencing
\end{IEEEkeywords}

@@

The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either interesting or boring. After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do. In \cite{idm}, another paper proposing using the LLM as a policy influencer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed. Here, too, the LLM response is fed back into a task-specific RL model, which in turn influences policy probabilities of certain actions being taken. In general, many of the novel methods suggested involve the LLM suggesting actions for the agent to perform in a given context, and its output influencing the agent policy.

@@

We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***. All methods use the same hyperparameters for the PPO, which can be found in the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one. We do this eight times per method, and compare the average results against each other.

@@

% We also notice a significant performance increase when using LLM policy influencing, even somewhat more than from reward shaping. Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM reward shaping. The standard deviation when using LLM policy influencing is also similarly large to PPO and reward shaping (NUMBER). By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.

Using LLM policy influencing in the DoorKey environment has resulted in somewhat worse results than using the PPO baseline. Averaging per episode over all eight runs shows that LLM policy influencing only achieves $89\%$ of the reward amount per episode that the baseline achieves. Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on. The method also gives the highest standard deviation of reward given per episode of the three compared so far, at $0.25$. As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.

***TODO: add paragraph about using both together*** 
% However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together. Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy influencing, with a similar standard deviation of NUMBER.

@@

Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER). LLM policy influencing gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping. Also here giving the highest average reward per episode, the combination of reward shaping and policy influencing gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy influencing.

\section{Discussion, Future Work and Conclusion}

% It works man, dogski did it!

% As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so. This has been done both through the LLM shaping the RL reward function, and through the LLM influencing the action probabilities of the RL agent, as well as through combining these two methodologies.

% For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing. We have implemented algorithms to perform this on top of a PPO baseline, and have measured varying results across environments of different complexities.

Our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment. A possible explanation for this could lie in the details of how the algorithms impact choice of action. LLM reward shaping does not directly impact decision making at all; it only impacts the reward function, meaning the mechanism that dictates which action is taken is still solely the actor network present in PPO. This means that reward shaping does not introduce any risk of the agent choosing a suboptimal action directly due to being told to do so by the LLM. Learning which action to take is still done through the same process of trial and error as in classical RL.

However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process. Every time the LLM suggests performing a suboptimal action, this impacts the action probabilities, regardless whether the PPO baseline has learned that another action is better or not. Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own. In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.

A natural suggestion for how to improve the performance of LLM policy influencing in the future thus becomes to decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses. One way of doing this is through annealing of the LLM values added to the logits. This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training. We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.
